Different forms of uncertainty differently affect the evolution of social learning

### 2022-10-18 ms.tex
Updates from Overleaf

In our model we allow just one trait to evolve, social learning; 
all other parameters are exogenous. Our primary outcome measure is the observed
frequency that the social learning trait fixates in simulated populations. We study
how the social learning fixation frequency responds to the four varieties of
uncertainty considered in this paper. \mt{I changed this somewhat, please double
check it's clear}

@@

\paragraph{Stopping conditions} The evolutionary process continues until the
population reaches evolutionary fixation with respect to social learning, where
$\sum_i s_i = N$ or $\sum_i s_i = 0$.

@@

A shorter effective lifespan, $L$, also favors the evolution of social learning in most cases. When effective lifespans are long and there are many opportunities to gather information by learning asocially, the value of social transmission is diminished. Given enough opportunities, all agents will learn the optimal behavior eventually. Thus, under most conditions, longer effective lifespans lead to a narrower range of the parameters $u$ and $B$ under which social learning was favored. It is noteworthy that in many cases, a large increase in the selection regime for social learning could be observed when $L=1$, and by definition no individual learning occurred. The exceptions to this general pattern occur for high levels of payoff ambiguity, $\pilow$ combined with low rates of environmental change, $u$ (left parts of plots on bottom row of Figure~\ref{fig:mainResults}). This reversal is partially explained by higher levels of drift when there is little difference in the payoffs between behaviors, and by the fact that the quality of social information is poorer when payoffs are ambiguous and lifespans are short (more on these effects of payoff ambiguity below).  
% \cm{is this right? I thought we should say something about the exceptions in this paragraph though we describe in more detail below, but not sure I have the right explanation}

Of the four types of uncertainty studied in this paper, payoff ambiguity had the most
complex effect on the evolution of social learning. Recall that we operationalized
payoff ambiguity with $\pilow$, the expected payoff for non-optimal behaviors
(equivalently the probability that those behaviors yielded a payoff of 1). This was
relative to $\pihigh = 0.9$, the expected payoff for the optimal behavior. When
payoff ambiguity was low, non-optimal behaviors rarely paid off, so that exploration
yielded reliable information and selection on learning strategies was  strong. When
payoff ambiguity is very high ($\pilow = 0.8$), two things happen. First, agents are
more likely to err by ascribing high value to non-optimal behaviors, since it is
more difficult for them to discern the difference between optimal and sub optimal
behaviors. Second, and perhaps more importantly, natural selection on strategies
that $do$ reliably select the optimal behavior is relatively weak. This means that
the effect of drift is relatively stronger, weakening selection in favor of social
learning under small values of $u$ and weakening selection against social learning
under high values of $u$ (Figure~\ref{fig:mainResults}, bottom row). 
This effect was especially strong under other types of uncertainty, i.e., larger selection set sizes and especially shorter effective lifespans---particularly when $L=1$ where no individual learning occurred.

@@

 Despite the common claim that copying others makes sense when one is uncertain, we find that different forms of uncertainty can either promote, or impede, the evolution of social learning. By disambiguating various formalizations of uncertainty in a single computational model, we have developed a more nuanced theoretical understanding of how uncertainty affects the evolution of social learning.
%We reviewed and distilled several empirical and theoretical models in the literature along four uncertainty dimensions: temporal environmental uncertainty, selection set size, ambiguous payoffs, and effective lifespan.  
We reproduced the well-known results that the uncertainty derived from a temporally variable environment can limit the evolution of social learning if environmental change is simply too rapid for intergenerational transmission to be of value. Importantly, we also showed that this relationship is affected by other sources of uncertainty. 
%Our model reproduced classic predictions that social learning can yield higher payoffs than individual learning when uncertainty is sufficiently limited. 
We found that larger sets of behavioral options, shorter effective lifespans, and increased payoff ambiguity could all favor the evolution of social learning. However, these effects were not always straightforward. 
Increased payoff ambiguity and decreased effective lifespan afforded a greater role for drift, which is essentially
evolutionary uncertainty in a finite-size population. Not surprisingly, when the consequence of one's choice does not affect payoffs much, selection either for or against social learning is substantially weaker. More interestingly, selection is also weaker when the ability to learn asocially is fraught with uncertainty. Our model demonstrated that social learning can act as a scaffold for individual learning---they are not opposite strategies. We endowed agents with a biologically-plausible individual learning mechanism, softmax search. This enabled agents to recover from the potential cost of receiving outdated information. This in turn enabled the evolution of social learning even when social information was likely to be outdated.

@@

Our modeling results suggest, on the one hand, that society might be better to
ignore social information from previous generations since our natural world is
rapidly changing~\citep{IPCC2022}; on the other hand, humans have increasingly more
behavioral options (selection set size) with high levels of uncertainty about which behaviors are most beneficial (payoff ambiguity), and in this case our model predicts a greater prevalence of social learning.
In general, detailed theoretical models like this are likely to be critical for understanding how humans might adapt to an uncertain and rapidly changing world.
Such understanding could suggest plausibly beneficial interventions to mitigate existential threats~\citep{Moya2020,Jones2021}, especially in
the most vulnerable communities~\citep{McNamara2020}, and to capitalize on new behavioral opportunities such as transitioning to clean energy use and
production~\citep{NatureEnergyEditorialPromisesPremises2018,Brisbois2022}. However, to better identify
potential pathways to effective adaptation it may be necessary to adjust some of our modeling assumptions to
include both the cumulative nature of cultural evolution and group structure of interpersonal exchange, both
of which have been shown to affect the sustainability of novel adaptations~\citep{Derex2020,Centola2018,Smaldino2021a}.
\cm{should we draw out some implications? I find this difficult, but e.g. stressing environmental change may discourage learning from older gen? having more tech options available for dealing with climate change may encourage social learning? Frankly so many of these problems seem to come back to encouraging people to socially learn from the RIGHT people, not just socially or asocially learning.} \mt{I gave this a try, with some restructuring to put the note about social learning as a scaffold for individual learning in the above paragraph.}

@@

\section{Number of prospective teachers sensitivty analysis}

We found that the number of prospectve teachers had little effect on $\meansl$
over $u$. There was slightly more drift when $N_T = 2$ compared to our 
main text analysis, slightly less drift than our main analysis when $N_T = 20$; 
we also tested $N_T=10$, which showed no obvious difference from our main results
with $N_T = 5$. When $N_T = 2$, increased drift is most pronounced in the
$\meansl$ curves when $\pilow=0.8$ (Figure~\ref{fig:nteachersSensitivity}a, 
bottom row). The transition from $\meansl = 1$ to $\meansl = 0$ is less sharp
across $L$ values, with social learning evolution being suppressed beginning
at smaller values of $u$. By contrast, when $N_T = 20$, transitions from 
$\meansl = 1$ to $\meansl = 0$ are sharper across $L$ when $\pilow = 0.8$
(Figure~\ref{fig:nteachersSensitivity}c, bottom row). Otherwise, the main 
results were reproduced across all tested values $N_T=2,5,10,20$.

@@

\section{Softmax parameter sensitivity analysis}

We tested two additional settings of the softmax behavior selection greediness parameter,
$\beta=1,100$, which reproduced qualitative patterns in the evolution of social learning
observed in the main text where we used $\beta=10$.  
Softmax greediness is achieved formally by increasing the
probability mass for selecting behaviors that have paid off more frequently over that agent's
life history. In other words, greater $\beta$ increases the probabilty that the optimal
observed behavior is chosen over all the rest. If $\beta$ is too small, agents will fail to
fully exploit their knowledge about which behavior is optimal because they too frequently
explore less profitable alternatives.  If $\beta$ is too great the agents may prematurely
settle on a sub-optimal behavior because it paid off first.  In a fuller evolutionary
exploration of model parameters one might allow $\beta$ to co-evolve, through
mutation and selection, with the social learner trait.  
However, our sensitivity analysis suggests this would be unnecessary 
since our main conclusions of how social learning evolves under uncertainty remain unchanged for different
$\beta$ over three orders of magnitude.

@@

The main results were again qualitatively obtained again with $\beta=100$
(Figure~\ref{fig:softmaxSensitivity}b). 
In some cases, especially when $\pilow=0.1$, individual learning may be performing
better with $\beta=100$ compared to results in the main text ($\beta=10$), 
evidenced by a quicker decline of social learning prevalence over $u$ when
$L > 1$ (Figure~\ref{fig:softmaxSensitivity}b,
top row).

\vspace{-3em} \begin{figure} %\addtocounter{figure}{-1} 
  \centering
  \caption{Sensitivity analysis of the main results for the softmax parameter $\beta = 100$ and
  % \addtocounter{figure}{-1}
  $\beta=1$. Recall the main results were obtained with $\beta = 10$.}
  \label{fig:softmaxSensitivity} \vspace{2em}
  \begin{subfigure}{\textwidth}
	\caption{$\beta = 1$}
	\includegraphics[width=\textwidth]{Figures/supplement/sensitivity_tau=1.0/mainResultsPlots.pdf}
  \end{subfigure}
\end{figure}
\newpage
\begin{figure}
  \ContinuedFloat
  \begin{subfigure}{\textwidth}
	\caption{$\beta = 100$}
	\includegraphics[width=\textwidth]{Figures/supplement/sensitivity_tau=0.01/mainResultsPlots.pdf}
  \end{subfigure}
\end{figure}

>>>>

In our model we allow just one trait to evolve: social learning. All other parameters are exogenous. Our primary outcome measure is the observed
frequency at which the social learning trait fixates in simulated populations. We study
how the social learning fixation frequency responds to each of the four varieties of
uncertainty considered in this paper. \mt{I changed this somewhat, please double
check it's clear} \ps{I edited a couple of words but looks good.}

@@

\paragraph{Stopping conditions} The evolutionary process continues until the
population reaches evolutionary fixation with respect to social learning, that is, when
$\sum_i s_i = N$ or $\sum_i s_i = 0$.

@@

A shorter effective lifespan, $L$, also favors the evolution of social learning in most cases. When effective lifespans are long and there are many opportunities to gather information by learning asocially, the value of social transmission is diminished. Given enough opportunities, all agents will learn the optimal behavior eventually. Thus, under most conditions, longer effective lifespans lead to a narrower range of the parameters $u$ and $B$ under which social learning was favored. It is noteworthy that in many cases, a large increase in the selection regime for social learning is observed when $L=1$, and by definition no individual learning can occur. The exceptions to this general pattern occur for high levels of payoff ambiguity, $\pilow$ combined with low rates of environmental change, $u$ (left parts of plots on bottom row of Figure~\ref{fig:mainResults}). This reversal is partially explained by higher levels of drift when there is little difference in the payoffs between behaviors, and by the fact that the quality of social information is poorer when payoffs are ambiguous and lifespans are short (more on these effects of payoff ambiguity below).  
% \cm{is this right? I thought we should say something about the exceptions in this paragraph though we describe in more detail below, but not sure I have the right explanation}

Of the four types of uncertainty studied in this paper, payoff ambiguity had the most
complex effect on the evolution of social learning. Recall that we operationalized
payoff ambiguity with $\pilow$, the expected payoff for non-optimal behaviors
(equivalently the probability that those behaviors yielded a payoff of 1). This was
relative to $\pihigh = 0.9$, the expected payoff for the optimal behavior. When
payoff ambiguity was low, non-optimal behaviors rarely paid off, so that exploration
yielded reliable information and selection on learning strategies was  strong. When
payoff ambiguity is very high ($\pilow = 0.8$), two things happen. First, agents are
more likely to err by ascribing high value to non-optimal behaviors, since it is
more difficult for them to discern the difference between optimal and sub optimal
behaviors. Second, and perhaps more importantly, natural selection on strategies
that $do$ reliably select the optimal behavior is relatively weak. This means that
the effect of drift is relatively stronger, weakening selection in favor of social
learning under low values of $u$ and weakening selection against social learning
under high values of $u$ (Figure~\ref{fig:mainResults}, bottom row). 
This effect was especially strong under other types of uncertainty, i.e., larger selection set sizes and especially shorter effective lifespans---particularly when $L=1$ where no individual learning occurred.

@@

 Despite the common claim that copying others makes sense when one is uncertain, we find that different forms of uncertainty can either promote or impede the evolution of social learning. By disambiguating various formalizations of uncertainty in a single computational model, we have developed a more nuanced theoretical understanding of how uncertainty affects the evolution of social learning.
%We reviewed and distilled several empirical and theoretical models in the literature along four uncertainty dimensions: temporal environmental uncertainty, selection set size, ambiguous payoffs, and effective lifespan.  
We reproduced the well-known results that the uncertainty derived from a temporally variable environment can limit the evolution of social learning if environmental change is simply too rapid for intergenerational transmission to be of value. Importantly, we also showed that this relationship is affected by other sources of uncertainty. 
%Our model reproduced classic predictions that social learning can yield higher payoffs than individual learning when uncertainty is sufficiently limited. 
We found that larger sets of behavioral options, shorter effective lifespans, and increased payoff ambiguity could all favor the evolution of social learning. However, these effects were not always straightforward. 
Increased payoff ambiguity and decreased effective lifespan afforded a greater role for drift, which is essentially
evolutionary uncertainty in a finite-size population. Not surprisingly, when the consequence of one's choice does not affect payoffs much, selection either for or against social learning is substantially weaker. More interestingly, selection is also weaker when the ability to learn asocially is fraught with uncertainty. Our model demonstrated that social learning can act as a scaffold for individual learning---they are not opposite strategies. We endowed agents with a biologically-plausible individual learning mechanism, softmax search. This enabled agents to recover from the potential cost of receiving outdated information. This in turn enabled the evolution of social learning even when social information was likely to be outdated.

@@

Our modeling results suggest, on the one hand, that it might sometimes be better to
ignore social information from previous generations since our natural world is
rapidly changing~\citep{IPCC2022}; on the other hand, humans have increasingly more
behavioral options (selection set size) with high levels of uncertainty about which behaviors are most beneficial (payoff ambiguity), and in this case our model predicts a greater prevalence of social learning.
In general, detailed theoretical models like this are likely to be critical for understanding how humans might adapt to an uncertain and rapidly changing world.
Such understanding could suggest plausibly beneficial interventions to mitigate existential threats~\citep{Moya2020,Jones2021}, especially in
the most vulnerable communities~\citep{McNamara2020}, and to capitalize on new behavioral opportunities such as transitioning to clean energy use and
production~\citep{NatureEnergyEditorialPromisesPremises2018,Brisbois2022}. However, to better identify
potential pathways to effective adaptation it may be necessary to adjust some of our modeling assumptions to
include both the cumulative nature of cultural evolution and group structure of interpersonal exchange, both
of which have been shown to affect the sustainability of novel adaptations~\citep{Derex2020,Centola2018,Smaldino2021a}.
\cm{should we draw out some implications? I find this difficult, but e.g. stressing environmental change may discourage learning from older gen? having more tech options available for dealing with climate change may encourage social learning? Frankly so many of these problems seem to come back to encouraging people to socially learn from the RIGHT people, not just socially or asocially learning.} \mt{I gave this a try, with some restructuring to put the note about social learning as a scaffold for individual learning in the above paragraph.}

@@

\section{Number of prospective teachers sensitivtiy analysis}

We found that the number of prospective teachers had little effect on $\meansl$
over $u$. There was slightly more drift when $N_T = 2$ compared to our 
main text analysis, slightly less drift than our main analysis when $N_T = 20$; 
we also tested $N_T=10$, which showed no obvious difference from our main results
with $N_T = 5$. When $N_T = 2$, increased drift is most pronounced in the
$\meansl$ curves when $\pilow=0.8$ (Figure~\ref{fig:nteachersSensitivity}a, 
bottom row). The transition from $\meansl = 1$ to $\meansl = 0$ is less sharp
across $L$ values, with social learning evolution being suppressed beginning
at smaller values of $u$. By contrast, when $N_T = 20$, transitions from 
$\meansl = 1$ to $\meansl = 0$ are sharper across $L$ when $\pilow = 0.8$
(Figure~\ref{fig:nteachersSensitivity}c, bottom row). Otherwise, the main 
results were reproduced across all tested values $N_T=2,5,10,20$.

@@

\section{Robust patterns, interesting variation, with different agent greediness}

We tested two additional settings of the softmax behavior selection greediness parameter,
$\beta=1,100$, which reproduced qualitative patterns in the evolution of social learning
observed in the main text where we used $\beta=10$.  
Softmax greediness is achieved formally by increasing the
probability mass for selecting behaviors that have paid off more frequently over that agent's
life history. In other words, greater $\beta$ increases the probability that the optimal
observed behavior is chosen over all the rest. If $\beta$ is too small, agents will fail to
fully exploit their knowledge about which behavior is optimal because they too frequently
explore less profitable alternatives.  If $\beta$ is too great the agents may prematurely
settle on a sub-optimal behavior because it paid off first.  In a fuller evolutionary
exploration of model parameters one might allow $\beta$ to co-evolve, through
mutation and selection, with the social learner trait.  
However, our sensitivity analysis suggests this would be unnecessary 
since our main conclusions of how social learning evolves under uncertainty remain unchanged for different
$\beta$ over three orders of magnitude.

@@

\mt{This is the greedier sensitivity analysis---need to frame it that way, note that excessive greediness suppresses the usefulness of social learning.} The main results were obtained with $\beta=100$
(Figure~\ref{fig:softmaxSensitivity}b). 
In some cases, especially when $\pilow=0.1$, individual learning may be performing
better with $\beta=100$ compared to results in the main text ($\beta=10$), 
evidenced by a quicker decline of social learning prevalence over $u$ when
$L > 1$ (Figure~\ref{fig:softmaxSensitivity}b,
top row).

\vspace{-3em} \begin{figure} %\addtocounter{figure}{-1} 
  \centering
  \caption{Sensitivity analysis of the main results for the softmax parameter $\beta = 100$ and
  % \addtocounter{figure}{-1}
  $\beta=1$. Recall the main results were obtained with $\beta = 10$.}
  \label{fig:softmaxSensitivity} \vspace{2em}
  \begin{subfigure}{\textwidth}
	\caption{$\beta = 1$}
	\includegraphics[width=\textwidth]{Figures/supplement/sensitivity_tau=1.0/mainResultsPlots.pdf}
  \end{subfigure}
\end{figure}

\begin{figure}
  \ContinuedFloat
  \begin{subfigure}{\textwidth}
	\caption{$\beta = 100$}
	\includegraphics[width=\textwidth]{Figures/supplement/sensitivity_tau=0.01/mainResultsPlots.pdf}
  \end{subfigure}
\end{figure}
